Language models as probabilistic cognitive models

### 2023-11-21 03-paper/paper-LLM-probmod.tex
minor revisions throughout whole doc

\begin{abstract}
  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks.
  But targeted assessment with methods inspired by experimental psychology highlights dissimilarities between human behavior and LLM predictions.
  Going even a step further, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
  We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level).
  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, thus building PCM-like predictions from LLM-derived information, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
\end{abstract}

@@

% A major contribution of this work is methodological.
This paper introduces and systematically compare different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and address the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures.
We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores, determined by a winner-takes-all strategy to identify a single target response.
We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not. \mf{double-check formulation of results}

The paper is structured as follows.
Section~\ref{motivation} motivates the investigation in this paper, by enlarging on the importance of deriving distributional predictions for conditional-level forced-choice data, and further elaborating on the relevant differences between LLMs and PCMs.
Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though the theoretically-motivated RSA model is.
Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivation: Distributional predictions for condition-level forced-choice data}
\label{motivation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of the same memory task.
Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
Nevertheless, each experimental condition can be, and often is, instantiated with different items, variation among which is deemed less relevant to the research question at hand.

Data from human participants for experiments of this kind usually shows some variability between items, and also variability between participants.
This variability is commonly incorporated in statistical analyses as random stochastic variation, e.g., by using hierarchical (regression) models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}.
Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the PCM will naturally be set-up to predict data at the level of the relevant experimental manipulation, i.e., at the condition-level.
Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, the focus of cognitive modeling is to explain systematic variation between experimental conditions.
Therefore, a basic PCM is essentially characterized by a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each logically possible data observation $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
The difference between a theory-driven PCM and a standard, theory-neutral statistical model (like a regression model), which may also define a parameterized likelihood function of condition-level data, is that the PCM's likelihood function is defined in terms that are considered to be conceptually meaningful for the research domain at hand (see Section~\ref{sec:model-pred-from} for a concrete example).

PCMs, thus construed, differ markedly from LLMs.
LLMs make predictions about the likelihood of words given a preceding or surrounding text.
They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at a level of generalizable abstraction which is arguably necessary for increasing human understanding of the phenomenon in question \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}.
Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
Whether or not these predictions are deemed explanatory, the empirical question arises of whether any of these predictions are empirically adequate.
This empirical question can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.

Standard benchmark-testing on multiple-choice tasks usually uses a winner-take-all (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench} to define accuracy scores.
The WTA-strategy can also be used to derive probabilistic item-level and condition-level predictions from LLMs.
Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$, and each of which .
For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same answer category $R_{i}$, e.g., that $y_{k1}$ is always the designated \emph{target option} that is assumed to be the ``true'' or to-be-selected goal answer, and that the other choice options are equally assignable into meaningful categories of answer types.

The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  \mf{How to extend to masked language models?}
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
}
%
\begin{align*}
  % S\left( y_{ki}, x_{k} \right) =
  \text{S}_{ki}
  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The WTA-based approach selects answers that maximize the item-level score (with random tie-breaking).
Concretely, if $B_{k} = \arg \max_{i} \text{S}_{ki}$ is the set of all options that maximize the item-level score for item $k$, then the \emph{item-level prediction} for item $I_{k}$ of the WTA is:
%
\begin{align*}
  P_{\text{item}}^{\text{WTA}} =
  \begin{cases}
    \frac{1}{\card{B_{k}}} & \text{if } y_{ki} \in B_{k} \text{, and} \\
    0                     & \text{otherwise.}
  \end{cases}
\end{align*}
%
Notice that the item-level WTA-prediction is usually a degenerate probability distribution placing all probability mass on a single choice option.

The \emph{condition-level prediction} of the WTA approach is obtained by averaging over the item-level predictions for each answer category:
%
\begin{align*}
  P_{\text{cond}}^{\text{WTA}}\left(R_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{WTA}} \left(y_{ki} \right)\,.
\end{align*}
%
The \emph{WTA-based accuracy} is the probability $P_{\text{cond}}^{\text{WTA}}\left(R_{1} \right)$ of selecting the target option over all instances of the task.
This is the most prevalent measure of the quality of LLM predictions.

@@

The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternatives.
This procedure ignores information about relative differences between scores among choice options; information which may be conceptually and practically relevant.
To generalize to less extreme sampling scenarios, the WTA approach can be generalized to softmax sampling, where the \emph{item-level prediction} of choosing option $y_{ki}$ is:
%
\begin{align*}
P_{\text{item}}^{\text{SM}}\left ( y_{ki} \right ) \propto \expo \left (\alpha \ \text{S}_{ki} \right)\,.
\end{align*}
%
The \emph{condition-level prediction} can be defined analogously to the above by averaging over items:
\begin{align*}
  P_{\text{cond}}^{\text{SM}}\left(R_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
\end{align*}
As before, the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(R_{i} \right)$, is the average probability of choosing the designated target option $y_{k1}$.
For $\alpha \rightarrow \infty$, the softmax strategy converges to the WTA strategy, but for finite $\alpha$ the approaches differ.
This has interesting entailments for accuracy assessment in more engineering-oriented contexts, as well as for questions of exploring the ``human-likeness'' of LLM predictions.

Looking at accuracy assessment, WTA-based accuracy measures can even differ qualitatively from accuracy measures obtained from the more general softmax-based strategy.
Differences in accuracy measures depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
The example below shows that categorically different results are, in principle, possible when using WTA- or softmax based accuracy.
The example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
%
\begin{quote}
  % \ex \label{bsp:exmpl-diff-WTA-SM}
  \textbf{Example:}
  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
  The WTA-based accuracy is 0.8.
  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
  As only the best option will be chosen, that information is irrelevant.
  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
\end{quote}
%
The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.

@@

As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
%
\begin{align*}
P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
\end{align*}
%
Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, are obtained from GPT-3.5 \mf{which model exactly, when assessed, cite paper}.
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
Allowing for random errors with probability $\epsilon_{C}$, as before,:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%
Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
    Approaches differ in the kind of item-level information they take into account and in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
  }
  \label{fig:measures-overview}
\end{figure}

@@

>>>>

\begin{abstract}
  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
  This raises questions about the human-likeness of LLM-derived information and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
  We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, thus building PCM-like predictions from LLM-derived information, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
\end{abstract}

@@

This paper introduces and systematically compares different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and addresses the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures.
We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores.
We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.

The paper is structured as follows.
Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way  LLMs and PCMs make predictions about data from forced-choice tasks.
Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions adequately capture the human data at the item-level.
Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Predictions for multiple-choice tasks from PCMs and LLMs}
\label{motivation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section argues that a major conceptual difference between LLMs and PCMs is that they predict data at different levels of abstraction: the item-level and the condition-level, as introduced below.
Another important difference is that PCMs are usually evaluated on a non-degenerate likelihood function, whereas common practice in LLM-evaluation is to use a winner-takes-all strategy.
Problems with this latter strategy are highlighted and alternative ways of defining item- and condition-level predictions from LLMs are introduced for subsequent empirical testing.

\paragraph{Item- and condition-level data.}
Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of the same memory task.
Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
Nevertheless, each experimental condition can be, and often is, instantiated with different items, variation among which is deemed less relevant to the research question at hand.

\paragraph{Probabilistic cognitive models.}
Data from human participants for experiments of this kind usually shows some variability between items, and also variability between participants.
This variability is commonly incorporated in statistical analyses as random stochastic variation, e.g., by using hierarchical (regression) models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}.
Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the PCM will naturally be set-up to predict data at the level of the relevant experimental manipulation, i.e., at the condition-level.
Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, the focus of cognitive modeling is to explain systematic variation between experimental conditions.
Therefore, a basic PCM is essentially characterized by a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each logically possible data observation $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
The difference between a theory-driven PCM and a standard, theory-neutral statistical model (like a regression model), which may also define a parameterized likelihood function of condition-level data, is that the PCM's likelihood function is defined in terms that are considered to be conceptually meaningful for the research domain at hand (see Section~\ref{sec:model-pred-from} for a concrete example).

\paragraph{PCMs vs.~LLMs.}
PCMs, thus construed, differ markedly from LLMs.
LLMs make predictions about the likelihood of words given a preceding or surrounding text.
They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at a level of generalizable abstraction which is arguably necessary for increasing human understanding of the phenomenon in question \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}.
Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
Whether any of these predictions are adequate is ultimately an empirical question, which can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.

\paragraph{Predictions from LLMs.}
Standard benchmark-testing on multiple-choice tasks usually uses a winner-take-all (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench} to define accuracy scores.
The WTA-strategy can also be used to derive probabilistic item-level and condition-level predictions from LLMs.
Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$, and each of which .
For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $r_{i}$ and that $y_{k1}$ always corresponds to the designated \emph{target option}, $r_{1}$, that is assumed to be the ``true'' or to-be-selected goal answer.

The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  \mf{TODO: Add quick word on how this can in principle be extended to masked language models.}
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
}
%
\begin{align*}
  % S\left( y_{ki}, x_{k} \right) =
  \text{S}_{ki}
  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The WTA-based approach selects answers that maximize the item-level score (with random tie-breaking).
Concretely, if $B_{k} = \arg \max_{i} \text{S}_{ki}$ is the set of all options that maximize the item-level score for item $k$, then the \emph{item-level prediction} for item $I_{k}$ of the WTA is:
%
\begin{align*}
  P_{\text{item}}^{\text{WTA}} =
  \begin{cases}
    \frac{1}{\card{B_{k}}} & \text{if } y_{ki} \in B_{k} \text{, and} \\
    0                     & \text{otherwise.}
  \end{cases}
\end{align*}
%
Notice that the item-level WTA-prediction is usually a degenerate probability distribution placing all probability mass on a single choice option.

The \emph{condition-level prediction} of the WTA approach is obtained by averaging over the item-level predictions for each response category:
%
\begin{align*}
  P_{\text{cond}}^{\text{WTA}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{WTA}} \left(y_{ki} \right)\,.
\end{align*}
%
The \emph{WTA-based accuracy} is the probability $P_{\text{cond}}^{\text{WTA}}\left(r_{1} \right)$ of selecting the target option $r_{1}$ over all instances of the task.
This is the most prevalent measure of the quality of LLM predictions.

@@

The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternatives.
This procedure ignores information about relative differences between scores among choice options; information which may be conceptually and practically relevant.
To generalize to less extreme sampling scenarios, the WTA approach can be generalized to softmax sampling, where the \emph{item-level prediction} of choosing option $y_{ki}$ is:
%
\begin{align*}
P_{\text{item}}^{\text{SM}}\left ( y_{ki} \right ) \propto \expo \left (\alpha \ \text{S}_{ki} \right)\,.
\end{align*}
%
The \emph{condition-level prediction} can be defined analogously to the above by averaging over items:
\begin{align*}
  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
\end{align*}
As before, the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, is the average probability of choosing the designated target response option $r_{1}$.

For $\alpha \rightarrow \infty$, the softmax strategy converges to the WTA strategy, but for finite $\alpha$ the approaches differ.
Indeed, WTA-based accuracy can even differ qualitatively from softmax-based accuracy, as shown by the following example.
The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
%
\begin{quote}
  % \ex \label{bsp:exmpl-diff-WTA-SM}
  \textbf{Example:}
  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
  The WTA-based accuracy is 0.8.
  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
  As only the best option will be chosen, that information is irrelevant.
  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
\end{quote}
%
The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.

@@

As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
%
\begin{align*}
P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
\end{align*}
%
Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, were obtained from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} .
% \footnote{
  % Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
The choice options are categorized as in the human experiment.\footnote{
  In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
Allowing for random errors with probability $\epsilon_{C}$, as before,:
%
\begin{align*}
  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
\end{align*}
%
The resulting likelihood function for item-level data with item-level predictions is:
%
\begin{align*}
 P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
\end{align*}
%
Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
    Approaches differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
  }
  \label{fig:measures-overview}
\end{figure}

@@

\begin{itemize}
  \item This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
\end{itemize}
